GRB Afterglows In The Multi-Messenger Era OR Electromagnetic Afterglows Of Structured Jets OR Structured Jets At All Angles

### 2019-08-05 structuredJets.tex
small updates

\afterglowpy{} uses the standard fourth order Runge-Kutta algorithm to numerically solve the $(R(t), u(t), \theta_j(t))$ system of ordinary differential equations on a fixed logarithmically spaced grid of $t$ \citep{Press:2007aa}.  The endpoints of the $t$-grid are chosen to bracket the burster frame times required to calculate the requested $\tobs$.  Initial conditions for $R$, $u$, and $\theta$ are typically those of a decelerating ultra-relativistic blast wave.  The user can set the density of the $t$-grid with the {\tt tRes} parameter: the number of grid points per decade of $t$.  The default value for {\tt tRes} is 1000, which is sufficiently dense that the blast wave ODE evolution is not the dominant error source but not so dense as to adversely impact performance.

@@

Each evaluation of the integrand requires a binary search to determine the burster time $t$ at which to evaluate $R$, $u$, and $\theta_j$.  Fluid quantities are then calculated using the shock-jump conditions and the synchrotron emissivity is evaluated using the standard external shock formulae \citep{Granot:2002aa,van-Eerten:2010aa}.

The numerical accuracy of a top-hat light curve with this integration scheme is typically better than $10^{-4}$.  The structured jet calculation splits the integration domain into {\tt latRes} disjoint annuli, each evolved as an independent top-hat and their emission summed.  By default {\tt latRes} is chosen such that there will be 5 zones per $\thC$-sized interval of $\theta$.   The choice of  {\tt latRes} has the largest impact on the code performance and accuracy.  We find the choice $\mathtt{latRes} = 5$ gives sufficiently quick performance at acceptable errors, typically on the order of $10^{-2}$.  TODO SHOW CONVERGENCE PLOT

@@

Using the definition of $\thC$ in Equation \eqref{eq:thCdef} we may write the angular energy profile near the jet axis as:
\begin{equation}
	E(\theta) = E_0 \left (1 - \frac{\theta^2}{2\thC^2} \right) + \OO(\theta^3) \label{eq:EthetaNear}
\end{equation}

The light curves of structured jets show two modes of behaviour, depending on whether the observer is \emph{aligned} ($\thobs < \thC$) or \emph{misaligned} ($\thobs > \thC$).
In the \emph{aligned} case, the light curves follow the standard on-axis top hat behaviour modified slightly for non-zero viewing angle.  There is little to distinguish between different structure $E(\theta)$, and the light curve is well-approximated by a broken power law with characteristic break times \citep{Granot:2002aa}.  The \emph{misaligned} light curve also may present as a broken power law, but with closure relations explicitly dependent on viewing angle and jet angular structure.

@@

\begin{figure}
	\includegraphics[width=\columnwidth]{figs/lc_on.pdf}
	\caption{Aligned ($\thobs = \thC/2$) afterglow light curves for top-hat (blue), Gaussian (orange), and power law (green) jets. When viewed on-axis, jets containing angular structure are very difficult to distinguish from top-hats.  The transition phase marking the onset of the jet break begins at $\tbm\sim3$ days with a shallow break and ends sharply at $\tbp \sim 100$ days.  These particular light curves use $\thC = 0.1$ rad, $\thW = 0.4$ rad, $E_0 = 10^{52}$erg, $n_0=10^{-2}$cm$^{-3}$, $p=2.2$, $\epse = 10^{-1}$, $\epsB = 10^{-3}$, and $d_L=40$Mpc. \label{fig:onaxis}}
\end{figure}

@@

\emph{Far off-axis} --- If the viewer lies outside the truncation angle, $\thobs > \thW$, then at early times the entire blast wave surface is off-axis and all emission is beaming suppressed.  This greatly reduces the early flux and leads to a dim but steeply rising light curve as the blast wave decelerates.  Emission is dominated by material on the edge nearest the observer and the angular size of the visible patch is constant in time.  This phase ends at $\tW$ when $\gamma^{-1} \sim \thobs-\thW$.  If $\thobs < \thW$ this phase is entirely absent.  Table \ref{tab:slopes} gives the temporal power law slope of the light curve in this phase.  These slopes are derived in Appendix \ref{app:derive1}.

@@

\begin{deluxetable*}{CCCCCCC}
	\tablecaption{Structured jet temporal and spectral slopes: $F_\nu \propto \tobs^{-\alpha} \nu^{-\beta}$. See Equations \eqref{eq:gdef} and \eqref{eq:geff} for the definition and effective values of $g$. \label{tab:slopes}}
	\tablehead{\colhead{Regime}& \colhead{Label} & \colhead{$\alpha_{\mathrm{off-axis}}$}& \colhead{$\alpha_{\mathrm{struct}}$}& \colhead{$\alpha_{\mathrm{struct,max}}$}& \colhead{$\alpha_{\mathrm{post, analytic}}$} & \colhead{$\beta$}}
	\startdata
	\nu<\nu_m<\nu_c     & D & 7 			& \frac{1 + 3g}{8+g} 		& 3		& -1/4		& 1/3 \\
	\nu_m<\nu<\nu_c     & G & 15/2 - 3p/2 	& \frac{3 - 6p +3g}{8+g} 	& 3		& -3p/4		& (1-p)/2 \\
	\nu < \nu_c < \nu_m & E & 17/3 		& \frac{-5/3 + 11g/3}{8+g} 	& 11/3	& -7/12		& 1/3 \\
	\nu_c < \nu < \nu_m & F & 13/2 		& \frac{-5 + 2g}{8+g} 		& 2		& -1			& -1/2 \\
	\nu_c < \nu_m < \nu & H & 8 - 3p/2 		& \frac{1-6p  +2g}{8+g} 	& 2		& -(3p+1)/4	& -p/2 \\ 
	\enddata
\end{deluxetable*}

@@

The $g$ parameter evolves with time as $\theta^*$ sweeps from the jet edge to the core.  This produces deviations in the light curve from a pure power law.  However, we find ultimately these deviations are not too large and the average slope is well approximated by $\geff  = g(\thobs/2)$.  For the power law and Gaussian jet models this gives:
\begin{align}
	\theff &= \frac{\thobs}{2}&& \text{Gaussian}\ , \\
	\theff &= \frac{\thobs}{\sqrt{1.8+2.1b^{-1.25} + (0.49-0.86b^{-1.25})\thobs/\thC}}&& \text{power law}\ . \label{eq:theff}
\end{align}
\begin{align}
	\geff &= \frac{\thobs^2}{4\thC^2} && \text{Gaussian}\ , \\
	\geff &= \frac{b \thobs^2}{4 \thC^2+\thobs^2} && \text{power law}\ . \label{eq:geff}
\end{align}
These values may be used in Table \ref{tab:slopes} for an approximate power law model of a structured jet with an appropriate viewing angle dependent temporal evolution.  These analytic results are in agreement with previous work on structured jets (CITE ROSSI) and the trans-relativistic numerical model described in Section \ref{sec:numerical}.

@@

\begin{figure}
	\includegraphics[width=\columnwidth]{figs/lc_off1.pdf}
	\caption{As Figure \ref{fig:onaxis} but $\thobs = 2\thC$.  The top-hat jet (with opening angle $\thC$) begins in the off-axis phase.  The Gaussian and power law jets begin in the structured phase as $\thobs < \thW$.  All three models enter similar post-jet break phases. \label{fig:offaxis1}}
\end{figure}
\begin{figure}
	\includegraphics[width=\columnwidth]{figs/lc_off2.pdf}
	\caption{As Figure \ref{fig:onaxis} but $\thobs = 6\thC$. All three models exhibit off-axis, structured, and post-jet break phases of evolution.  At larger opening angles the difference between structured models is more evident.  \label{fig:offaxis2}}
\end{figure}

>>>>

\afterglowpy{} uses the standard fourth order Runge-Kutta algorithm to numerically solve the $(R(t), u(t), \theta_j(t))$ system of ordinary differential equations (Equations \eqref{eq:ODE1}-\eqref{eq:ODE3}) on a fixed logarithmically spaced grid of $t$ \citep{Press:2007aa}.  The endpoints of the $t$-grid are chosen to bracket the burster frame times required to calculate the requested $\tobs$.  Initial conditions for $R$, $u$, and $\theta$ are typically those of a decelerating ultra-relativistic blast wave.  The user can set the density of the $t$-grid with the {\tt tRes} parameter: the number of grid points per decade of $t$.  The default value for {\tt tRes} is 1000, which is sufficiently dense that the blast wave ODE evolution is not the dominant error source but not so dense as to adversely impact performance.

@@

Each evaluation of the integrand requires a binary search to determine the burster time $t$ at which to evaluate $R$, $u$, and $\theta_j$.  Fluid quantities are then calculated using the shock-jump conditions and the synchrotron emissivity is evaluated using the standard external shock formulae (Equation \eqref{eq:syncSpec}, see also \citet{Granot:2002aa,van-Eerten:2010aa}).

The numerical accuracy of a top-hat light curve with this integration scheme is typically better than $10^{-4}$.  The structured jet calculation splits the integration domain into {\tt latRes} disjoint annuli, each evolved as an independent top-hat and their emission summed.  By default {\tt latRes} is chosen such that there will be 5 zones per $\thC$-sized interval of $\theta$.   The choice of  {\tt latRes} has the largest impact on the code performance and accuracy.  We find the choice $\mathtt{latRes} = 5$ gives sufficiently quick performance at acceptable errors, typically on the order of $10^{-2}$.

@@

Previous work has demonstrated the light curves of structured jets show two modes of behaviour, depending on whether the observer is \emph{aligned} ($\thobs < \thC$) or \emph{misaligned} ($\thobs > \thC$) \citep{Rossi:2002aa, Granot:2002ab, Panaitescu:2003aa, van-Eerten:2010aa}.
In the \emph{aligned} case, the light curves follow the standard on-axis top hat behaviour modified slightly for non-zero viewing angle.  There is little to distinguish between different structure $E(\theta)$, and the light curve is well-approximated by a broken power law with characteristic break times \citep{Granot:2002aa}.  The \emph{misaligned} light curve structure is more complicated.  It also presents as a broken power law, but with closure relations explicitly dependent on viewing angle and jet angular structure.

We can write the received flux as:
\begin{equation}
	F_\nu = I_\nu^*\ \Delta \Omega\ , 
\end{equation}
where $I_\nu^* = I_\nu(\theta^*)$ is the peak specific intensity, located at $\theta=\theta^*$ and $\phi=0$, and $\Delta \Omega$ is the effective solid angle of the blast wave contributing to the emission.

From the definition of $\thC$ in Equation \eqref{eq:thCdef} we can see $E(\theta) = E_0(1 - \theta^2/(2\thC^2) + \OO(\theta^3))$.

@@

%\begin{figure}
%	\includegraphics[width=\columnwidth]{figs/lc_on.pdf}
%	\caption{Aligned ($\thobs = \thC/2$) afterglow light curves for top-hat (blue), Gaussian (orange), and power law (green) jets. When viewed on-axis, jets containing angular structure are very difficult to distinguish from top-hats.  The transition phase marking the onset of the jet break begins at $\tbm\sim3$ days with a shallow break and ends sharply at $\tbp \sim 100$ days.  These particular light curves use $\thC = 0.1$ rad, $\thW = 0.4$ rad, $E_0 = 10^{52}$erg, $n_0=10^{-2}$cm$^{-3}$, $p=2.2$, $\epse = 10^{-1}$, $\epsB = 10^{-3}$, and $d_L=40$Mpc. \label{fig:onaxis}}
%\end{figure}

@@

\emph{Far off-axis} --- If the viewer lies outside the truncation angle, $\thobs > \thW$, then at early times the entire blast wave surface is off-axis and all emission is beaming suppressed.  This greatly reduces the early flux and leads to a dim but steeply rising light curve as the blast wave decelerates.  Emission is dominated by material on the edge nearest the observer and the angular size of the visible patch is constant in time.  This phase ends at $\tW$ when $\gamma^{-1}(\thW) \sim \thobs-\thW$.  If $\thobs < \thW$ this phase is entirely absent.  Table \ref{tab:slopes} gives the temporal power law slope of the light curve in this phase.  These slopes are derived in Appendix \ref{app:derive1}.

@@

\begin{deluxetable*}{CCCCCCCC}
	\tablecaption{Structured jet closure relations: $F_\nu \propto \tobs^{\alpha} \nuobs^{\beta}$. The ``far-off-axis'' phase occurs at early times and only if there is no material directly in the observer's line of sight ($\thobs > \thW$). The ``generic'' slope applies whenever the jet is relativistic, non-spreading, and the observer is within the beaming cone of some part of the jet ($|\thobs - \theta_*| < \gamma^{-1}(\theta_*)$). The $g$ parameter is positive and depends on the viewing angle $\thobs$ and the specific structure profile $E(\theta)$.  See Equations \eqref{eq:gdef} and \eqref{eq:geff} for the definition and effective values of $g$.  For a top hat jet $g=0$ and for a Gaussian jet $g \approx \thobs^2/(4\thC^2)$. The $\som$ parameter encodes how the effective angular size of the jet is changing with Lorentz factor: $\Delta \Omega \propto \gamma^{-\som}$.  The ``structured'' phase is seen by misaligned viewers ($\thobs > \thC$), occurs before the jet break, and corresponds to $\som \approx 1$. The ``pre-jet break'' phase is seen by aligned viewers ($\thobs < \thC$).  The ``post-jet break'' slope given here does not include the effects of jet spreading and as such is an over-estimate of the true post-jet break slope. \label{tab:slopes}}
	\tablehead{\colhead{Regime}& \colhead{Label} & \colhead{$\alpha_{\mathrm{far-off-axis}}$} & \colhead{$\alpha_{\mathrm{generic}}$} & \colhead{$\alpha_{\mathrm{struct}}$}& \colhead{$\alpha_{\mathrm{pre}}$}& \colhead{$\alpha_{\mathrm{post, no-spreading}}$} & \colhead{$\beta$}}
	\startdata
	& & & & (\som = 1) & (\som=2, g=0) & (\som=0, g=0) & \\[0.3cm]
	\nu<\nu_m<\nu_c     & D	& 7 		& \dfrac{-2 + 3\som + 3g}{8+g} 		& \dfrac{1 + 3g}{8+g} 	&1/2		& -1/4	& 1/3 \\[0.3cm]
	\nu < \nu_c < \nu_m & E	& 17/3 	& \dfrac{-14/3 + 3\som + 11g/3}{8+g}  & \dfrac{-5/3 + 11g/3}{8+g} &1/6 	& -7/12	& 1/3 \\[0.3cm]
	\nu_c < \nu < \nu_m & F	& 13/2 	& \dfrac{-8 + 3\som + 2g}{8+g} 		& \dfrac{-5 + 2g}{8+g} 	& -1/4 	& -1		& -1/2 \\[0.3cm]
	\nu_m<\nu<\nu_c     & G 	&15/2 - 3p/2 & \dfrac{-6p + 3\som + 3g}{8+g} 	& \dfrac{3 - 6p +3g}{8+g} 	& -3(p-1)/4 & -3p/4	& (1-p)/2 \\[0.3cm]
	\nu_m, \nu_c < \nu & H 	& 8 - 3p/2 	& \dfrac{-6p - 2 + 3\som + 2g}{8+g} 	& \dfrac{1-6p  +2g}{8+g} 	& -3p/4	& -(3p+1)/4 & -p/2 \\[0.3cm]
	\enddata
\end{deluxetable*}

@@

The $g$ parameter evolves with time as $\theta^*$ sweeps from the jet edge to the core.  This produces deviations in the light curve from a pure power law.  However, we find ultimately these deviations are not too large and the average slope is well approximated by $\geff  = g(\thobs/2)$.  For the power law and Gaussian jet models this gives:
\begin{align}
	\theff &= \frac{\thobs}{2}&& \text{Gaussian}\ , \\
	\theff &= \frac{\thobs}{\sqrt{1.8+2.1b^{-1.25} + (0.49-0.86b^{-1.25})\thobs/\thC}}&& \text{power law}\ . \label{eq:theff}
\end{align}
\begin{align}
	\geff &= \frac{\thobs^2}{4\thC^2} && \text{Gaussian}\ , \\
	\geff &= \frac{2 b (\thobs-\theff)\theff}{b \thC^2+\theff^2} && \text{power law}\ . \label{eq:geff}
\end{align}
These values may be used in Table \ref{tab:slopes} for an approximate power law model of a structured jet with an appropriate viewing angle dependent temporal evolution.  These analytic results are in agreement with previous work on structured jets (CITE ROSSI) and the trans-relativistic numerical model described in Section \ref{sec:numerical}.

@@

%\begin{figure}
%	\includegraphics[width=\columnwidth]{figs/lc_off1.pdf}
%	\caption{As Figure \ref{fig:onaxis} but $\thobs = 2\thC$.  The top-hat jet (with opening angle $\thC$) begins in the off-axis phase.  The Gaussian and power law jets begin in the structured phase as $\thobs < \thW$.  All three models enter similar post-jet break phases. \label{fig:offaxis1}}
%\end{figure}
%\begin{figure}
%	\includegraphics[width=\columnwidth]{figs/lc_off2.pdf}
%	\caption{As Figure \ref{fig:onaxis} but $\thobs = 6\thC$. All three models exhibit off-axis, structured, and post-jet break phases of evolution.  At larger opening angles the difference between structured models is more evident.  \label{fig:offaxis2}}
%\end{figure}
